---
title: "Assignment 3_Claxton_funds of ML"
output: pdf_document
date: "2026-06-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

library(caret)
library(reshape)
library(e1071)
library(gmodels)
```
```{r}
bank.df<-read.csv("C:\\Users\\User\\OneDrive\\Desktop\\Grad School_Business Analytics\\Fundamentals of ML\\Assignment 3\\UniversalBank-1.csv")
#read CSV and save as dataframe

bank.df <- bank.df[, c("Online", "CreditCard", "Personal.Loan")] 
#reduce variables to Online, CC, and Personal.Loan

bank.df$Online <- factor(bank.df $Online, levels = c(0,1)) 

bank.df$CreditCard     <- factor(bank.df $CreditCard,     levels = c(0,1)) 

bank.df$Personal.Loan   <- factor(bank.df $Personal.Loan,   levels = c(0,1)) 
#convert to factors

set.seed(123)
#set seed

```

```{r}
# Stratified partition: 60% training 

train_index <- createDataPartition(bank.df $Personal.Loan, p = 0.6, list = FALSE) 

train <- bank.df [train_index, ] 

valid <- bank.df [-train_index, ] 

train$dummy <- 1 

train_melt <- melt(train, 

                   id.vars = c("CreditCard", "Online", "Personal.Loan"), 

                   measure.vars = "dummy") 
#create data partition, establish train and test
#use melt() to convert data into long form
pivot_A <- cast(train_melt, 
  CreditCard + Personal.Loan ~ Online, 
  fun.aggregate = sum, 
  value = "value")
#Use case to create the pivot tale. 
#Aggregate and value are just counting rows
pivot_A
```
Part A: This is the pivot table with credit card and personal loan on the rows with online as a column variable
Part B: Consider a customer who owns a credit card and uses online banking. Looking at the pivot table, the probability of this customer accepting a private loan (based on the training data) is 51/(467+51) = 0.098.

```{r}
# (Part C) Pivot tables for Loan as a function of Online 

pivot_LoansByOnline <- table(train$Personal.Loan, train$Online) 

pivot_LoansByOnline 

# Pivot table for loan as a function of CC 

pivot_LoansByCC <- table(train$Personal.Loan, train$CreditCard) 

pivot_LoansByCC 
```
Part C: These two pivot tables depict (from top to bottom) Loan as a function of Online status and Loan as a function of Credit Card status.

Part D:
P(CreditCard = 1 | Loan = 1) = 0.29
P(Online=1 | Loan=1) = 0.61
P(Loan = 1) 0.10
P(CreditCard=1 |Loan = 0) = 0.29
P(Online=1 | Loan = 0) = .59
P(Loan=0) = .90

Part E: Compute the Naive Bayes probability = Probability of accepting a loan when the user has a credit card and uses online banking -> P(Loan=1 |CreditCard=1, Online=1) -> Loan =1  / Loan=0 + Loan = 1 -> .29 * .61 * .1 / ((.29 * .61 * .1)+(.29 * .59 * .90)) = 0.103

Part G: What components do we need? We need all of the components of the table because we need the probabilities to compute Naive Bayes
```{r}
nb_model <- naiveBayes(Personal.Loan ~ CreditCard + Online, data = train) 

nb_model 

#Conditional probabilities for Online | Personal.Loan and CC
# | Personal.Loan should match the answers from part D 

new_customer <- data.frame( 

  CreditCard = factor(1, levels = c(0,1)), 

  Online = factor(1, levels = c(0,1))) 

#create the new customer 

pred_probs <- predict(nb_model, new_customer, type = "raw") 

pred_probs 

#predict the class of the new customer 
#using the naiveBayes model 

#This should match part E and NOT part B 
```
This matches our calculations for part E because we manually calculated probabilities using Naive Bayes methodology. It does *NOT match exactly to part B because those probabilities were derived from empirical observations from the training data. 
